---
title: "Getting Started with szKendall"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with szKendall}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The goal of szKendall is to measure the dissimilarity between two single cells based on Hi-C data, accounting for the discrepancies in structural zero positions. This rank-based dissimilarity measure can be powerful for single-cell Hi-C clustering and other downstream analyses. The calculation is implemented using Rcpp for speed and supports parallel computation via `foreach` and `doParallel` packages.

## Installation

You can install the szKendall package from GitHub:

```{r, warning=FALSE, message=FALSE, results='hide'}
# If not already installed:
# install.packages("devtools")

# library(devtools)
# devtools::install_github("https://github.com/osu-stat-gen/szKendall")
suppressPackageStartupMessages({
  library(ggplot2)
  library(plsgenomics)
  library(dplyr)
  library(Rtsne)
  library(umap)
  library(szKendall)
})
```

You may also need to load parallel backends:

```{r setup, warning=FALSE, message=FALSE}
suppressPackageStartupMessages({
library(doParallel)
library(foreach)
})
registerDoParallel(cores = 4)  
# Run foreach::registerDoSEQ() to override any parallel backend and force sequential calculation 
# foreach::registerDoSEQ()
```

The `szKendall()` function supports three input formats:

1. A **\#LP × \#cells** numeric matrix 

2. A **list** of length **\#cells**, where each element is a symmetric Hi-C count matrix of dimension **\#LP X \#LP** 

3. A **3D array** of dimension **\#LP X \#LP X \#cells**, where each slice corresponds to a symmetric Hi-C count matrix.

The argument `dist.method` specifies the weighting scheme and accepts either  
`"szKendall"` or `"szKendall1"`.  

See `?szKendall` for further details.

## Example: Calculating szKendall Dissimilarity and Perform Clustering

This is an example which shows you how to calculate the szKendall dissimilarity between cells, where contact is a built-in data input in the third format supported by the package.
If `heatmap = TRUE`, the function will automatically generate heatmaps of the resulting distance matrix. The default setting is FALSE.

```{r, fig.width=12, fig.height=4}
szK  <- szKendall(contacts=contacts,dist.method="szKendall",heatmap=TRUE)
```


```{r}
szK1 <- szKendall(contacts=contacts,dist.method="szKendall1")
```


`scHiC_hm_dist()` also provides convenient visualization tools for the dissimilarity matrices.

```{r viz-heatmap, fig.width=12, fig.height=4}
par(mfrow=c(1,3))
scHiC_hm_dist(szK1[[1]]) # szKendall(1)

```


The `szKendall` package additionally offers 2D visualization tools, such as t-SNE (`scHiC_tSNE_dist()`) and UMAP (`scHiC_UMAP_dist()`) projections of the dissimilarity matrix.

```{r viz-2dscatter, fig.width=5, fig.height=5}
# Specify the true cell labels
n_single <- 50
cell.type <- factor(c(rep(1,n_single), rep(2,n_single), rep(3,n_single)), levels=c("1", "2", "3"))

# t-SNE 2D projection
scHiC_tSNE_dist(szK[[1]], cell_type=cell.type, seed=1234, title="szK", alpha_value=1)

# UMAP 2D projection
scHiC_UMAP_dist(szK[[1]], cell_type=cell.type, seed=1234, main="szK", alpha_value=1)
```

## Computing four assessment criteria

We can assess clustering performance using four criterion values (ARI, WSS%, ASW, and MIS). Below is an example.

```{r clustering}
suppressWarnings({
  suppressPackageStartupMessages({
    library(cluster)  # required to perform PAM clustering
    library(factoextra)  # required to perform hierarchical clustering
  })
})


##  Perform PAM clustering for number of cluster from 1 to 10 and plot the results 
set.seed(1234)
pam2 <- pam(szK[[1]], k=3, nstart=10, diss=TRUE)
# ARI 
ARI(cell.type, pam2$clustering)  
# WSS% 
WSS_percentage(pam2$clustering, szK[[1]])
# ASW 
# (the ASW extracted from 'pam2$silinfo$avg.width' should be identical to the one returned by 'ASW' function.) 
ASW(pam2$clustering, szK[[1]])
pam2$silinfo$avg.width
# MIS 
MIS(pam2$clustering, szK[[1]])



## Simulate single-cell Hi-C Data 

In this package, we also provide a function that simulates the single-cell Hi-C data for `n_single` cells of the same cell (sub)type, given the 3D coordinates of `n` loci. The underlying expected contact count data are also returned, which indicate all the structural zero positions (i.e., those positions with expected count being 0). 

```{r simulation} 
# Simulate the data for 50 cells
n_single <- 50
# Load the 3D coordinates of n=61 loci
data("coord3D_T1")

# Save the resulting data as ST1 (subtype 1), which contains the simulated and expected contact counts 
# (Both matrices are of dimension 1830 by 50, where 1830 = 61*60/2 is the number of locus pairs, 
#  and 50 is the number of cells (n_single). )
set.seed(1234)
ST1 <- scHiC_simulate_withoutdiagonal(
  data = coord3D_T1,
  alpha_0 = 1.8,
  alpha_1 = -1,
  beta_l = 0.9,
  beta_g = 0.9,
  beta_m = 0.9,
  gamma = 0.6,
  tau1 = 0.7,
  eta = 0.1,
  tau2=0.6,
  n_single=n_single
)
dim(ST1$singledat)
dim(ST1$truecount)
```


### Explanation of Parameters

* `data` is an `n` by 3 matrix representing the 3D coordinates of `n` loci.
* `alpha_0` controls the sequencing depth of the simulated single-cell Hi-C data.
* `alpha_1` reflects the biophysical law between the expected contact counts and the 3D Euclidean distances among the loci. Default value is -1.
* `beta_l` mimics the effect of fragment length in generating single-cell Hi-C data. Default value is 0.9.
* `beta_g` mimics the effect of GC content in generating single-cell Hi-C data. Default value is 0.9.
* `beta_m` mimics the effect of mappability in generating single-cell Hi-C data. Default value is 0.9.
* `gamma` defines the threshold for expected contact counts; locus pairs with expected counts below this percentile could become structural zero candidates.
* `tau1` sets the probability that a locus pair with expected count below `gamma` percentile turning into a structural zero candidate.
* `eta` decides the proportion of common structural zeros among the structural zero candidates.
* `tau2` decides the proportion of cell-specific structural zeros among the remaining `1 - eta` proportion of the structural zero candidates.
* `n_single` is the number of single cells (i.e., columns) to be generated.


### Explanation of Output

`scHiC_simulate_withoutdiagonal()` returns a list with 2 elements:

* `singledat`: The simulated single-cell Hi-C contact counts. The dimension is `N` by `n_single`, where `N = n*(n-1)/2` is the number of unique locus pairs given `n` loci.
* `truecount`: The expected ("true") single-cell Hi-C contact counts. Same dimension as `singledat`.


